Four decades of water quality change in the upper San Francisco Estuary

### 2017-01-09 sftrends_manu.Rnw
annual per chg removed from trndmap

<<obsdat, fig.cap = 'Observed \\ac{din} from ten stations in the upper \\ac{sfe} Delta.  Data were collected monthly and evaluated with \\ac{wrtds} models from 1976 to 2013. Note different y-axis scales.  See \\cref{fig:delt_map} for station locations.', fig.height = 8, fig.width = 6, cache = T, out.width = '0.8\\textwidth'>>=
data(delt_dat)

@@

# combine all
trnds_seasyr <- rbind(trnds_nrm, trnds_fit, trnds_obs) %>% 
  select(Site_Code, resvar, cat, ztest, tau, perchg, ann, trndvar) %>% 
  filter(!trndvar %in% 'res') %>% 
  mutate(
    Site_Code = factor(Site_Code, levels = c('D7', 'D6', 'D4', 'D28', 'D26', 'D19', 'P8', 'C10', 'C3')),
    ztest = ifelse(ztest < 0.05, '(< 0.05)', '(ns)'),
    resvar = factor(resvar, levels = reslabs$shrt, labels = reslabs$expr), 
    trndvar = factor(trndvar, levels = c('fits', 'norm'), labels = c('Pred.', 'Flow-norm.'))
  ) %>% 
  unite('trnd', trndvar, ztest, sep = ' ')

@@

# year only
toplo1 <- filter(trnds_seasyr, cat %in% c('1976-1995', '1996-2013'))
p1 <- ggplot(toplo1, aes(x = tau, y = Site_Code)) + 
  geom_point(aes(pch = trnd, colour = abs(tau)), size = 4, alpha = 0.8) + 
  facet_grid(cat~resvar, labeller = label_parsed
    ) + 
  geom_vline(xintercept = 0, linetype = 'dashed') + 
  mythm + 
  xlab(expression(tau)) +
  scale_shape_manual(values = c(16, 1, 17, 2)) +
  scale_colour_gradient(low = 'grey60', high = 'black', guide = F) + 
  scale_x_continuous(limits = c(-1, 1)) + 
  geom_hline(yintercept = c(3.5, 6.5), colour = 'grey', alpha = 0.6) + 
  scale_y_discrete('Stations')

# seasonal, first years
toplo2 <- filter(trnds_seasyr, !cat %in% c('1976-1995', '1996-2013')& ann == 'bef') 
p2 <- ggplot(toplo2, aes(x = tau, y = Site_Code)) + 
  geom_point(aes(pch = trnd, colour = abs(tau)), size = 4, alpha = 0.8) + 
  facet_grid(cat~resvar, labeller = label_parsed
    ) + 
  geom_vline(xintercept = 0, linetype = 'dashed') + 
  mythm + 
  xlab(expression(tau)) +
  scale_shape_manual(values = c(16, 1, 17, 2)) +
  scale_colour_gradient(low = 'grey60', high = 'black', guide = F) + 
  scale_x_continuous(limits = c(-1, 1)) + 
  geom_hline(yintercept = c(3.5, 6.5), colour = 'grey', alpha = 0.6) + 
  scale_y_discrete('Stations')

# seasonal, first years
toplo3 <- filter(trnds_seasyr, !cat %in% c('1976-1995', '1996-2013')& ann == 'aft')
p3 <- ggplot(toplo3, aes(x = tau, y = Site_Code)) + 
  geom_point(aes(pch = trnd, colour = abs(tau)), size = 4, alpha = 0.8) + 
  facet_grid(cat~resvar, labeller = label_parsed
    ) + 
  geom_vline(xintercept = 0, linetype = 'dashed') + 
  mythm + 
  xlab(expression(tau)) +
  scale_shape_manual(values = c(16, 1, 17, 2)) +
  scale_colour_gradient(low = 'grey60', high = 'black', guide = F) + 
  scale_x_continuous(limits = c(-1, 1)) + 
  geom_hline(yintercept = c(3.5, 6.5), colour = 'grey', alpha = 0.6) + 
  scale_y_discrete('Stations')

@@

<<trndmap, fig.cap = 'Percent change per year in nitrogen analytes for aggregations by (a) year, (b) seasons from 1976-1995, and (c) seasons from 1996-2013.  Changes are based on seasonal Kendall tests of flow-normalized results within each time period. Station names are shown in the top left panel. Values less than -9\\% change were truncated for visual clarity of the bar plots, see \\cref{tab:trndsann,tab:trndsbef,tab:trndsaft} for actual values. Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.', fig.height = 8, fig.width = 8, cache = T, out.width = '\\textwidth'>>=

@@

# unnest data frame, assign colors by rank and sign
toplo <- select(trnds_seasyr, Site_Code, resvar, ann, cat, norm) %>% 
  rename(perchg = norm) %>% 
  left_join(., statmeta, by = 'Site_Code') %>% 
  data.frame %>% 
  mutate(
    perchg = pmax(yrng[1], perchg),
    perchg = pmin(yrng[2], perchg),
    negs = sign(perchg),  
    rnks = scales::rescale(abs(perchg), c(1, 1000)), 
    rnks = round(rnks), 
    cols = ifelse(negs == 1, palpos[rnks], palneg[rnks]),
    ind = 1:nrow(.)
    ) %>%
  unite('cattype', cat, ann, sep = '', remove = F) %>%
  mutate(
    cattype = factor(cattype),
    cattype = fct_collapse(cattype, 
      an = c('1976-1995bef', '1996-2013aft'),
      mobef = paste0(c('Spring', 'Summer', 'Fall', 'Winter'), 'bef'), 
      moaft = paste0(c('Spring', 'Summer', 'Fall', 'Winter'), 'aft') 
    )
  )

@@

  # add site labels if plot is first
  pall <- pbase
  set.seed(5)
  if(which(uni == unis) == 1)
    pall <- pbase + 
      geom_label_repel(
        data = statmeta, 
        aes(x = Longitude, y = Latitude, label = Site_Code), 
        label.r = unit(0, "lines"),
        box.padding = unit(2, "lines"),
        label.padding = unit(0.15, "lines"), 
        point.padding = unit(0.1, "lines"),
        force = 2, size = 2,
        fill = 'grey90'
      )

@@

grid.arrange(
  leg,
  arrangeGrob(
    arrangeGrob (plo_ls[[1]], plo_ls[[4]], plo_ls[[7]], ncol = 1, top = c('(a) 1976-1995, 1996-2013')), 
    arrangeGrob(plo_ls[[3]], plo_ls[[6]], plo_ls[[9]], ncol = 1, top = c('(b) Sp, Su, Fa, Wi (1976-1995)')),
    arrangeGrob(plo_ls[[2]], plo_ls[[5]], plo_ls[[8]], ncol = 1, top = c('(c) Sp, Su, Fa, Wi (1996-2013)')),
    ncol = 3
    ),
  ncol = 1, 
  heights = c(0.1, 1)
  )

@@

ptheme <- theme_minimal() + 
  theme(
    axis.line.x = element_line(), 
    axis.line.y = element_line(), 
    axis.ticks.x = element_line(),
    axis.ticks.y = element_line(),
    axis.ticks.length = unit(.1, "cm"),
    axis.title.x = element_blank()
  )
lncol <- RColorBrewer::brewer.pal(9, 'Set1')[2]

@@

p1 <- ggplot(toplo2) + 
  geom_area(aes(x = date, y = val, group = var, fill = var), stat = 'identity', position = 'stack') + 
  scale_x_date(expand = c(0, 0), limits = xrng) + 
  scale_fill_brewer(palette = 'Blues', labels = leglabs) + 
  geom_vline(xintercept = as.numeric(lndate), linetype = 'dashed') +
  geom_label_repel(
    data = data.frame(x = lndate, y = 35), 
    aes(x, y, label = 'WWTP upgrades'), 
    box.padding = grid::unit(2, "lines"), 
    nudge_x = 1
    ) +
  scale_y_continuous(ylab1, expand = c(0, 0)) + 
  ptheme + 
  theme(
    legend.position = 'top', 
    legend.title = element_blank()
    )

>>>>

<<obsdat, fig.cap = 'Observed \\ac{din} (NH$_4^{+}$ + NO$_2^{-}$/NO$_3^{2-}$) from ten stations in the upper \\ac{sfe} Delta.  Data were collected monthly and evaluated with \\ac{wrtds} models from 1976 to 2013. Note different y-axis scales.  See \\cref{fig:delt_map} for station locations.', fig.height = 7, fig.width = 6, cache = T, out.width = '0.9\\textwidth'>>=
data(delt_dat)

@@

# combine all
trnds_seasyr <- rbind(trnds_nrm, trnds_fit, trnds_obs) %>% 
  select(Site_Code, resvar, cat, ztest, tau, perchg, ann, trndvar) %>% 
  filter(!trndvar %in% 'fits') %>% 
  mutate(
    Site_Code = factor(Site_Code, levels = c('D7', 'D6', 'D4', 'D28', 'D26', 'D19', 'P8', 'C10', 'C3')),
    ztest = ifelse(ztest < 0.05, '(p < 0.05)', '(ns)'),
    resvar = factor(resvar, levels = reslabs$shrt, labels = reslabs$expr), 
    trndvar = factor(trndvar, levels = c('res', 'norm'), labels = c('Obs.', 'Flow-norm.'))
  ) %>% 
  unite('trnd', trndvar, ztest, sep = ' ')

@@

# year only
toplo1 <- filter(trnds_seasyr, cat %in% c('1976-1995', '1996-2013'))
p1 <- ggplot(toplo1, aes(x = tau, y = Site_Code)) + 
  geom_point(aes(pch = trnd, colour = abs(tau)), size = 4, alpha = 0.8) + 
  facet_grid(cat~resvar, labeller = label_parsed
    ) + 
  geom_vline(xintercept = 0, linetype = 'dashed') + 
  mythm + 
  xlab(expression(tau)) +
  scale_shape_manual(values = c(1, 16, 2, 17)) +
  scale_colour_gradient(low = 'grey60', high = 'black', guide = F) + 
  scale_x_continuous(limits = c(-1, 1)) + 
  geom_hline(yintercept = c(3.5, 6.5), colour = 'grey', alpha = 0.6) + 
  scale_y_discrete('Stations')

# seasonal, first years
toplo2 <- filter(trnds_seasyr, !cat %in% c('1976-1995', '1996-2013')& ann == 'bef') 
p2 <- ggplot(toplo2, aes(x = tau, y = Site_Code)) + 
  geom_point(aes(pch = trnd, colour = abs(tau)), size = 4, alpha = 0.8) + 
  facet_grid(cat~resvar, labeller = label_parsed
    ) + 
  geom_vline(xintercept = 0, linetype = 'dashed') + 
  mythm + 
  xlab(expression(tau)) +
  scale_shape_manual(values = c(1, 16, 2, 17)) +
  scale_colour_gradient(low = 'grey60', high = 'black', guide = F) + 
  scale_x_continuous(limits = c(-1, 1)) + 
  geom_hline(yintercept = c(3.5, 6.5), colour = 'grey', alpha = 0.6) + 
  scale_y_discrete('Stations')

# seasonal, first years
toplo3 <- filter(trnds_seasyr, !cat %in% c('1976-1995', '1996-2013')& ann == 'aft')
p3 <- ggplot(toplo3, aes(x = tau, y = Site_Code)) + 
  geom_point(aes(pch = trnd, colour = abs(tau)), size = 4, alpha = 0.8) + 
  facet_grid(cat~resvar, labeller = label_parsed
    ) + 
  geom_vline(xintercept = 0, linetype = 'dashed') + 
  mythm + 
  xlab(expression(tau)) +
  scale_shape_manual(values = c(1, 16, 2, 17)) +
  scale_colour_gradient(low = 'grey60', high = 'black', guide = F) + 
  scale_x_continuous(limits = c(-1, 1)) + 
  geom_hline(yintercept = c(3.5, 6.5), colour = 'grey', alpha = 0.6) + 
  scale_y_discrete('Stations')

@@

<<trndmap, fig.cap = 'Percent change per year in nitrogen analytes for aggregations by (a) year, (b) seasons from 1976-1995, and (c) seasons from 1996-2013.  Changes are based on seasonal Kendall tests of flow-normalized results within each time period. Station names are shown in the top left panel. Values less than -9\\% change were truncated for visual clarity of the bar plots, see \\cref{tab:trndsann,tab:trndsbef,tab:trndsaft} for actual values. Months for each season are Spring: MAM, Summer: JJA, Fall: SON, Winter: DJF.', fig.height = 8, fig.width = 5.3, cache = T, out.width = '0.7\\textwidth'>>=

@@

# unnest data frame, assign colors by rank and sign
toplo <- select(trnds_seasyr, Site_Code, resvar, ann, cat, norm) %>% 
  rename(perchg = norm) %>% 
  left_join(., statmeta, by = 'Site_Code') %>% 
  data.frame %>% 
  mutate(
    perchg = pmax(yrng[1], perchg),
    perchg = pmin(yrng[2], perchg),
    negs = sign(perchg),  
    rnks = scales::rescale(abs(perchg), c(1, 1000)), 
    rnks = round(rnks), 
    cols = ifelse(negs == 1, palpos[rnks], palneg[rnks]),
    ind = 1:nrow(.)
    ) %>%
  filter(!cat %in% c('1976-1995', '1996-2013')) %>% # remove year categories
  unite('cattype', cat, ann, sep = '', remove = F) %>%
  mutate(
    cattype = factor(cattype),
    cattype = fct_collapse(cattype, 
      mobef = paste0(c('Spring', 'Summer', 'Fall', 'Winter'), 'bef'), 
      moaft = paste0(c('Spring', 'Summer', 'Fall', 'Winter'), 'aft') 
    )
  )

@@

  # add site labels if plot is first
  pall <- pbase
  set.seed(5)
  if(which(uni == unis) == 2)
    pall <- pbase + 
      geom_label_repel(
        data = statmeta, 
        aes(x = Longitude, y = Latitude, label = Site_Code), 
        label.r = unit(0, "lines"),
        box.padding = unit(2, "lines"),
        label.padding = unit(0.15, "lines"), 
        point.padding = unit(0.1, "lines"),
        force = 2, size = 2,
        fill = 'grey95'
      )

@@

grid.arrange(
  leg,
  arrangeGrob(
    arrangeGrob(plo_ls[[2]], plo_ls[[4]], plo_ls[[6]], ncol = 1, top = c('(a) Sp, Su, Fa, Wi (1976-1995)')),
    arrangeGrob(plo_ls[[1]], plo_ls[[3]], plo_ls[[5]], ncol = 1, top = c('(b) Sp, Su, Fa, Wi (1996-2013)')),
    ncol = 2
    ),
  ncol = 1, 
  heights = c(0.1, 1)
  )

@@

ptheme <- theme_minimal() + 
  theme(
    axis.line.x = element_line(), 
    axis.line.y = element_line(), 
    axis.ticks.x = element_line(),
    axis.ticks.y = element_line(),
    axis.ticks.length = unit(.1, "cm"),
    axis.title.x = element_blank()
  )
cols <- RColorBrewer::brewer.pal(9, 'Blues')[c(3, 5, 8)]

@@

p1 <- ggplot(toplo2) + 
  geom_area(aes(x = date, y = val, group = var, fill = var), stat = 'identity', position = 'stack') + 
  scale_x_date(expand = c(0, 0), limits = xrng) + 
  scale_fill_manual(values = cols, labels = leglabs) + 
  geom_vline(xintercept = as.numeric(lndate), linetype = 'dashed') +
  geom_label_repel(
    data = data.frame(x = lndate, y = 35), 
    aes(x, y, label = 'WWTP upgrades'), 
    box.padding = grid::unit(2, "lines"), 
    nudge_x = 1
    ) +
  scale_y_continuous(ylab1, expand = c(0, 0)) + 
  ptheme + 
  theme(
    legend.position = 'top', 
    legend.title = element_blank()
    )
